---
title: "Analysing Phytochemical Diversity - An introduction to *chemdiv*"
author: "Hampus Petrén, et al"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analysing Phytochemical Diversity - An introduction to *chemdiv*}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

**The *chemdiv* package and this vignette are in development and may not be fully functional/complete**

## Introduction
*chemdiv* is an R package used to analyse phytochemical data. It contains
functions centered around calculations of phytochemical diversity and
dissimilarity. This includes the use of diversity and dissimilarity indices 
that incorporate the biosynthetic and/or structural properties of the 
phytochemical compounds for the calculations, thereby providing comprehensive 
measures of phytochemical diversity.

*chemdiv* is currently available on GitHub: <https://github.com/hpetren/chemdiv>

This introduction serves as a tutorial explaining the intended use of all the
functions in the package.


## Setup
The developmental version of the package can be installed directly from GitHub
in R using the `install_github` function from the `devtools` package.

```{r, eval = FALSE}
# Install and load devtools package
install.packages("devtools")
library("devtools")

# Install chemdiv
install_github("hpetren/chemdiv")
```

```{r}
# Load chemdiv
library(chemdiv)
```


## Using *chemdiv*

## Data formatting
Analyses in *chemdiv* will work best for sets of data, commonly generated 
by chemical ecologists using GC-MS, LC-MS or similar, where all or most 
compounds in the samples have been confidently identified.

Two separate datasets are needed to use the full set of analyses included in 
the package. We illustrate this by using an example dataset, that is
included in the package, on floral scent of X individuals from Y populations 
for the plant *Arabis alpina*.
---Link to whatever dataset is used---


The first dataset, which we call the *proportions* dataset, should be a 
data frame containing data on the relative proportions of different 
compounds (columns) in different samples (rows):
```{r}
data("alpinaDataSubset")
head(alpinaDataSubset)[,1:5]
```
The dataset contains the proportional concentration of different floral
scent compounds in different samples. The first column indicates which
population each sample belongs to. For convenience, we separate this column
from the rest of the dataset:
```{r}
alpinaPopulation <- alpinaDataSubset$Population
alpinaData <- alpinaDataSubset[,-1]
```

The second dataset, which we call the *compounds* dataset, should be a 
data frame containing, in each of three columns in a data frame,
the common name, SMILES and InChIKey IDs of all the compounds that are 
present in the first dataset:
```{r}
data("alpinaCompSubset")
head(alpinaCompSubset)
```
SMILES and InChIKey are chemical identifiers that are easily obtained for 
each compound by searching for it in 
[PubChem](https://pubchem.ncbi.nlm.nih.gov/),
or by using various automated tools such as the 
[PubChem Identifier Exchange Service](https://pubchem.ncbi.nlm.nih.gov/idexchange/idexchange.cgi) 
or [The Chemical Translation Service](https://cts.fiehnlab.ucdavis.edu/).
The user is intentionally required to do this manually to ensure correctness,
as lists of compounds very often contain compounds wrongly named, 
wrongly formatted, under various synonyms etc. which prevents easy automatic 
translation of compound names to SMILES and InChIKey.

The name and order of the compounds in the first dataset (column names) has 
to be the same as the name and order of the compounds in the second dataset
(compound column): ---Double check if this is really necessary---
```{r}
table(colnames(alpinaData) == alpinaCompSubset$compound) # Should be all TRUE
```

Once data is correctly formatted we can begin with analyses. 

### Compound classification and dissimilarities
Two functions are used to classify and compare phytochemical compounds,
and are applied to the *compounds* dataset.

The function `NPCTable()` classifies compounds with 
[NPClassifier](https://npclassifier.ucsd.edu/), and puts the results in a 
dataframe containing the pathway, superclass and class for each compound.

```{r}
alpinaNPC <- NPCTable(alpinaCompSubset)
alpinaNPC[1,]
```
NPClassifier tries to classify each compound into a hierarchical classification,
largely corresponding to their biosynthesis, of pathway, superclass and class.
This classification can be used by other *chemdiv* functions.


### Diversity calculations

### Sample dissimilarities

### Molecular networks

### Chemodiversity and network plots
Plot, network, shortcut







