---
title: "Analysing Phytochemical Diversity - An introduction to *chemdiv*"
author: "Hampus Petrén, et al"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analysing Phytochemical Diversity - An introduction to *chemdiv*}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

**The *chemdiv* package and this vignette are in development and may not be fully functional/complete**

## Introduction
*chemdiv* is an R package used to analyse phytochemical data. It contains
functions centered around calculations of phytochemical diversity and
dissimilarity. This includes the use of diversity and dissimilarity indices 
that incorporate the biosynthetic and/or structural properties of the 
phytochemical compounds for the calculations, thereby providing comprehensive 
measures of phytochemical diversity.

*chemdiv* is currently available on GitHub: <https://github.com/hpetren/chemdiv>

This introduction serves as a tutorial explaining the intended use of all the
functions in the package.


## Setup
The developmental version of the package can be installed directly from GitHub
in R using the `install_github` function from the `devtools` package.

```{r, eval = FALSE}
# Install and load devtools package
install.packages("devtools")
library("devtools")

# Install chemdiv
install_github("hpetren/chemdiv")
```

```{r}
# Load chemdiv
library(chemdiv)
```


## Using *chemdiv*

### Data formatting
Analyses in *chemdiv* will work best for sets of data, commonly generated 
by chemical ecologists using GC-MS, LC-MS or similar, where all or most 
compounds in the samples have been confidently identified.

Two separate datasets are needed to use the full set of analyses included in 
the package. We illustrate this by using an example dataset, that is
included in the package, on floral scent of X individuals from Y populations 
for the plant *Arabis alpina*.
---Link to whatever dataset is used---


The first dataset, which we call the *proportions* dataset, should be a 
data frame containing data on the relative proportions of different 
compounds (columns) in different samples (rows):
```{r}
data("alpinaDataSubset")
head(alpinaDataSubset)[,1:5]
```
The dataset contains the proportional concentration of different floral
scent compounds in different samples. The first column indicates which
population each sample belongs to. For convenience, we separate this column
from the rest of the dataset:
```{r}
alpinaPopulation <- alpinaDataSubset$Population
alpinaData <- alpinaDataSubset[,-1]
```

The second dataset, which we call the *compounds* dataset, should be a 
data frame containing, in each of three columns in a data frame,
the common name, SMILES and InChIKey IDs of all the compounds that are 
present in the first dataset:
```{r}
data("alpinaCompSubset")
head(alpinaCompSubset)
```
SMILES and InChIKey are chemical identifiers that are easily obtained for 
each compound by searching for it in 
[PubChem](https://pubchem.ncbi.nlm.nih.gov/),
or by using various automated tools such as the 
[PubChem Identifier Exchange Service](https://pubchem.ncbi.nlm.nih.gov/idexchange/idexchange.cgi) 
or [The Chemical Translation Service](https://cts.fiehnlab.ucdavis.edu/).
The user is intentionally required to do this manually to ensure correctness,
as lists of compounds very often contain compounds wrongly named, 
wrongly formatted, under various synonyms etc. which prevents easy automatic 
translation of compound names to SMILES and InChIKey.

The name and order of the compounds in the first dataset (column names) has 
to be the same as the name and order of the compounds in the second dataset
(compound column): ---Double check if this is really necessary---
```{r}
table(colnames(alpinaData) == alpinaCompSubset$compound) # Should be all TRUE
```

Once data is correctly formatted we can begin with analyses. 

### Compound classification and dissimilarities
Two functions are used to classify and compare phytochemical compounds,
and are applied to the *compounds* dataset.

The function `NPCTable()` classifies compounds with 
[NPClassifier](https://npclassifier.ucsd.edu/) (Kim et al. 2021), and puts 
the results in a dataframe containing the pathway, superclass and class 
for each compound.

```{r}
alpinaNPC <- NPCTable(compoundData = alpinaCompSubset)
alpinaNPC[1,] # Classification of first compound in dataset
```
NPClassifier tries to classify each compound into a hierarchical classification,
largely corresponding to their biosynthesis, of pathway, superclass and class.
This classification can be used by other *chemdiv* functions.

The function `compDis()` calculates pairwise dissimilarities between
phytochemical compounds. Dissimilarity calculations can be based on
the biosynthesis and/or structure of the compounds. `type = "NPClassifier"`
calculates dissimilarities based on the classification by NPClassifier.
`type = "PubChemFingerprint"` and `type = "fMCS"` calculates dissimilarities
based on the structure of the compounds with two different methods.

We calculate compound dissimilarities with `type = "PubChemFingerprint"`.

```{r}
alpinaCompDis <- compDis(compoundData = alpinaCompSubset,
                         type = "PubChemFingerprint")

alpinaCompDis$fingerDisMat[1:4, 1:4] # Part of compound dissimilarity matrix
```
The output from the function is a list with one or several compound 
dissimilarity matrices, depending on which type(s) was used as input. 
This matrix is can be visualized and used by the other functions in the package.


### Diversity calculations
Phytochemical diversity can be calculated with functions `calcDiv()`, 
`calcBetaDiv()` and `calcDivProf()`.

`calcDiv()` calculates alpha-diversity for each sample (row). The function
can calculate a number of different diversity and evenness indices.
We recommend calculating diversity as Hill numbers (Chao et al. 2014),
as this provides a number of advantages, including the use of the parameter 
q which controls the sensitivity of the measure to the relative 
abundance of compounds. This can be done as normal Hill diversity, 
which depends on compound richness and evenness, and as functional 
Hill diversity,which additionally depends compound dissimilarity 
(Chiu & Chao 2014), utilizing the compound dissimilarity matrix 
from `compDis()`.

We calculate functional Hill diversity for q = 1. 
```{r}
alpinaDiv <- calcDiv(sampleData = alpinaData, 
                     compDisMat = alpinaCompDis$fingerDisMat,
                     type = "FuncHillDiv",
                     q = 1)

head(alpinaDiv)
```
The function outputs a data frame with the selected type(s) of measures
as columns.

***

`calcDivProf()` can be used to calculate normal or functional Hill diversity
for a range of q-values simultaneously, generating a so called diversity
profile.

We calculate a diversity profile for functional Hill diversity, using the
default range of q = 0 to q = 3. 
```{r}
alpinaDivProf <- calcDivProf(sampleData = alpinaData,
                             compDisMat = alpinaCompDis$fingerDisMat,
                             type = "FuncHillDiv")

head(alpinaDivProf$divProf)[,1:5] # Part of the diversity profile data frame
```
The function outputs a list with input parameters and the diversity profile
as a data frame.

***

`calcBetaDiv()` calculates beta-diversity for a set of samples, in the
Hill number framework. The function calculates a single beta-diversity value 
for the supplied sample data. This is calculated as beta = gamma / alpha.
```{r}
alpinaBetaDiv <- calcBetaDiv(sampleData = alpinaData,
                             compDisMat = alpinaCompDis$fingerDisMat,
                             type = "FuncHillDiv")

alpinaBetaDiv
```
The output from the function is a list with one or several dissimilarity
matrices, depending on which type(s) was used as input. This matrix is
can be visualized and used by the other functions in the package.


### Sample dissimilarities
Phytochemical dissimilarity can be calculated with the function `sampleDis()`.
This calculates Bray-Curtis dissimilarities and/or Generalized UniFrac 
dissimilarities between samples, utilizing the compound dissimilarity 
matrix from `compDis()`.

We calculate phytochemical dissimilarity as Generalized UniFrac dissimilarities:
```{r, message = FALSE}
alpinaSampleDis <- sampleDis(sampleData = alpinaData,
                             compDisMat = alpinaCompDis$fingerDisMat,
                             type = "GenUniFrac")

alpinaSampleDis$GenUniFrac[1:4, 1:4] # Part of sample dissimilarity matrix
```
The output from the function is a list with one or several sample dissimilarity
matrices, depending on which type(s) was used as input.


### Molecular network
Function `molNet()` uses a matrix generated by the `compDis()` function to 
create a molecular network of the phytochemical compounds, and calculates 
some properties of the network.

We create a molecular network based on the compound dissimilarity matrix.
We can also include the the NPClassifier classification. We manually set
`cutOff = 0.75` in this case.
```{r}
alpinaNetwork <- molNet(compDisMat = alpinaCompDis$fingerDisMat,
                        npcTable = alpinaNPC,
                        cutOff = 0.75)

summary(alpinaNetwork)
```
The output is a list including the network, and some basic network parameters.


### Chemodiversity and network plots
There are two functions that can be used to conveniently plot the
molecular network and different types phytochemical diversity and 
dissimilarity calculated by the above functions.

`molNetPlot()` creates a basic plot of the molecular network generated
by the molNet function. We create a single molecular network for the whole
dataset, including compound names and the classification by NPClassifier.
```{r, fig.width = 12, fig.height = 8, out.width = "90%"}
molNetPlot(sampleData = alpinaData,
           networkObject = alpinaNetwork$networkObject,
           npcTable = alpinaNPC,
           plotNames = TRUE)
```

Nodes are compounds. Nodes are coloured by their "pathway" classification
by NPClassifier. Node size represents the mean proportional concentration of
the compounds. Edge widths represent compound similarity, and are only plotted
for similarities higher than the cutoff value. 

***

`chemDivPlot()` can be used to conveniently create basic plots of the 
different types of chemodiversity measurements created by functions above.
Four types of plots can be created. With argument `compDisMat` a 
dendrogram visualizing compound dissimilarities is created. With argument 
`divData`, diversity/evenness is visualized with a boxplot.
With argument `divProfData` a diversity profile will be created. With argument `sampleDisMat` sample dissimilarities will be visualized as an NMDS plot.
Grouping data can be supplied with argument `groupData`.
```{r, fig.width = 12, fig.height = 8, out.width = "90%"}
chemDivPlot(compDisMat = alpinaCompDis$fingerDisMat,
            divData = alpinaDiv,
            divProfData = alpinaDivProf,
            sampleDisMat = alpinaSampleDis$GenUniFrac,
            groupData = alpinaPopulation)
```

The output consists of the selected plots. Any combination of plots
can be generated.


### Shortcut function
The function `quickChemDiv()` uses many of the above functions to in one 
simple step calculate/visualize chemodiversity for users wanting to 
quickly explore their data using standard parameters. This can be used
to generate the same four types of plots as above.

```{r, eval = FALSE}
quickChemDiv(compoundData = alpinaCompSubset,
             sampleData = alpinaData,
             groupData = alpinaPopulation,
             outputType = "plots") # Not run
```



## References

Paper for dataset

Paper for chemdiv eventually

Kim et al 2021

Chao et al. 2014

Chiu & Chao 2014



